Efficient online Sequential Monte Carlo smoother for partially observed stochastic differential equations

### 2017-08-10 TexSubmission/answer_referees.tex

\vspace{.3cm}

\item
\item
\item
\item
\item {\em Can anything be said about the bias of the proposed estimator for finite $N$? Is it
only due to the form of  $\Lambda_k^N$ ?}

\vspace{.3cm}

The bias of forward backward based particle smoothers due to $\Lambda_k^N$ (i.e. to the particle approximations) was analyzed in \cite{delmoral:doucet:singh:2010}. In the case of additive functional as in (2), the bias of the forward only version of the FFBS algorithm is shown to be upper bounded by a term of order $n/N$ for strongly mixing hidden Markov models.

@@

A CLT was derived for the original version of PaRIS algorithm as the number $N$ of particles grows to infinity. A CLT may also derived for our extension as it is a noisy version of the original method. However, even if its proof is not too involved it still require additional assumptions and technicalities. And the asymptotic variance remains hard to analyze (for instance to choose the parameters in comparison to other smoothing methods). Analyzing the variance of particle smoothers (FFBS, FFBSi, PaRIS, two-filer) is at the core of ongoing works and we believe that comparing the variance would be more suitable and interesting than just providing a CLT with no rules to choose between methods. We are also working on the estimation of such asymptotic variances which is a complex problem out of the scope of this paper.

\item {\em It is worth mentioning that, for the log-growth model, $X$ is a function of $\sigma$ and what
are the consequences of that in a context where $\sigma$ is an unknown parameter.}

\vspace{.3cm}

Even if there is no actual implementation of the EM algorithm in the paper for the log-growth model, we agree that providing some insights on the way it could be used may be very useful. We added a paragraph at the end of the log-growth section following the work of [4] explaining how the process can be transformed to estimate $\sigma$ using an EM algorithm although $X$ is a function of $\sigma$. \textcolor{red}{Pas si simple, la transformation en deux etapes de Beskos et al ne fonctionne pas comme le processus X n'est pas directement observe...}

@@

\item {\em page 16, line 49: For the proof to be valid, the induction result needs to be established for $k\ge 0$.}

\vspace{.3cm}

\end{enumerate}

\bibliographystyle{plain}
\begin{thebibliography}{10}

\bibitem{delmoral:doucet:singh:2010}
P.~{D}el {M}oral, {A}.~{D}oucet, and {S}.{S}.~{S}ingh.
\newblock A backward particle intepretation of {F}eynman-{K}ac formulae.
\newblock {\em {ESAIM:~M2AN}}, 44:947--975, 2010.
\end{thebibliography}

### 2017-08-10 TexSubmission/answer_referees.tex

{\em On page 15, line 13, I believe that $h(m)$ has sneaked its way into the equation and should be removed.}

\vspace{.3cm}
 $h(m)$ was removed in this equation.

\subsection*{Referee 2}

\subsubsection*{Major comments}
\begin{enumerate}
\item \textcolor{red}{Je pense que c'est le gros morceau... La reponse a cette question recouvre en fait les questions 1/4/5/6/8 donc quasiment tout ce qu'il reste.
On s'est focalises sur ces EDS essentiellement pour l'utilisation des estimateurs sans biais et leurs applications a des modeles en ecologie. C'est facile d'expliquer la partie gradient et d'utiliser Ait-Sahalia pour expliquer que c'est dur en non scalaire. Ensuite, je pense que le mieux c'est de bien decrire que les estimateurs de densites peuvent s'obtenir avec tous les EA et dans ce cas detailler les hypotheses proprement pour justifier notre approche. On peut aller jusqu'a l'EA3, qui sera quadratique probablement mais remplit la condition A3 donc peut etre a priori utilise. Ensuite, je pense que l'on peut mentionner que ce que l'on fait peut s'etendre a d'autres EDS avec une diffusion non unitaire grace aux tres recentes methodes d'importance sampling. La team Roberts n'a pas encore publie son algo (le fera-t-elle un jour ?) mais Kohatsu-Higa si et son algor ressemble beaucoup a l'approche Warwickienne. On a dans ce cas un estimateur sans biais, et que l'on peut simuler, sous des hypotheses plus generales. On ne change pas tout notre papier, on garde une focalisation de type Poisson mais on peut faire un paragraphe que se passe-t-il dans le cas non unitaire... PaRIS marche exactement pareil, a chacun d'utiliser son estimateur de densite prefere. What ya think ?
Je change dans cet esprit, vous me direz ce que vous en pensez !}

We agree with the Referee that the model was not introduced properly and that the use of generalized Poisson estimators required additional comments. This remark led to a major modification of the beginning of Section~2. The assumptions mentioned by the Referee are highlighted clearly and much more commented to explain the setting in which GPE can be used.

As some assumptions may be restrictive (specifically the gradient assumption on the drift), we also provide alternative to GPE to estimate transition densities. The only requirement to extend PaRIS algorithm is to obtain an unbiased estimator than can be evaluated explicitly to compute the importance weights. A recent class of estimators which do not use the gradient assumptions are the continuous importance sampling based methods of \cite{}. They introduce other sets of assumptions such as...

\item {\em When presenting equation (2), the authors should explain the importance of approximating those type of functionals.}

@@

\vspace{.3cm}

\end{enumerate}

\bibliographystyle{plain}
\begin{thebibliography}{10}

\bibitem{delmoral:doucet:singh:2010}
P.~{D}el {M}oral, {A}.~{D}oucet, and {S}.{S}.~{S}ingh.
\newblock A backward particle intepretation of {F}eynman-{K}ac formulae.
\newblock {\em {ESAIM:~M2AN}}, 44:947--975, 2010.

\bibitem{olsson:westerborn:2016}
J.~Olsson and J.~Westerborn.
\newblock Efficient particle-based online smoothing in general hidden {M}arkov
  models: the {PaRIS} algorithm.
\newblock {\em ArXiv:1412.7550}, 2016.

\end{thebibliography}

### 2017-08-10 TexSubmission/ParisEM.tex
Additional answers

\section{The Generalized Random PaRIS algorithm}
\label{sec:rwparis}
$(X_t)_{t\ge 0}$ is defined as a weak solution to the following SDE in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion. It is assumed that $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function. The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{0\le k \le n}$ in $(\mathbb{R}^m)^{n+1}$. 
For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot) = g_k(X_k)$. 
The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
For all $0\le k \le n-1$, the conditional distribution of $X_{k+1} $ given $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.

@@

\subsection*{Bounded estimator of $q_k$}
For $x, y \in \mathbb{R}^d$, by Girsanov and Ito's formulas, the transition density $q_k(x,y)$ of \eqref{eq:target:sde} satisfies, with $\Delta_k = t_{k+1}-t_k$,
\begin{align*}
q_k(x,y)=\varphi_{\Delta_k}(x,y)\exp\left\lbrace A(y)-A(x)\right\rbrace \mathbb{E}_{\mathbb{W}^{x,y,\Delta_k}}\left[ \exp \left\lbrace - \int_0^{\Delta_k} \phi(\mw_s)\rmd s \right\rbrace \right]\eqsp,
\end{align*}
where $\mathbb{W}^{x,y,\Delta_k}$ is the law of Brownian bridge starting at $x$ at 0 and hitting $y$ at $\Delta_k$, $(\mw_t)_{0\leq t \leq \Delta_k}$ is such a Brownian bridge, $\varphi_{\Delta_k}(x,y)$ is the p.d.f. of a normal distribution with mean $x$ and variance $\Delta_k$, evaluated at $y$ and $\phi:\mathbb{R}^d\to\mathbb{R}$ is defined as:
\[
\phi(x) =\left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2\eqsp,
\]
with $\triangle$ the Laplace operator.
Assume that there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta_k$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$.
Let $\kappa$ be a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and $(U_j)_{1\le j\le \kappa}$ be independent uniform random variables on $[0,\Delta_k]$, and $
\zeta_k = \left\{\kappa,\mw,U_1,\ldots,U_\kappa\right\}\eqsp$. 
As shown in \cite{fearnhead:papaspiliopoulos:roberts:2008}, a positive unbiased estimator is given by 
\begin{multline}
\widehat{q}_k(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)\right\}\\ 
\times\mathrm{exp}\left\{-\U_\mw\Delta\right\}\frac{\Delta_k^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\U_\mw-\phi(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
\end{multline}
Interesting choices of $\mu$ are discussed in \cite{fearnhead:papaspiliopoulos:roberts:2008} and we focus here on the so called GPE-1, where $\mu$ is a Poisson distribution with intensity $(\U_\mw-\Lo_\mw)\Delta_k$. In that case, the estimator \eqref{eq:unbiased:q} becomes:
\begin{equation}
\widehat{q}_{k}(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\prod_{j=1}^{\kappa}\frac{\U_\mw-\phi(\mw_{U_j})}{\U_\mw-\Lo_\mw}\eqsp.\label{eq:GPE1}
\end{equation}
On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to satisfy of the assumptions \eqref{assumpt:global}-\eqref{assumpt:local} is that the function:
\begin{align}
\rho_{\Delta_k}:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\nonumber\\
(x,y)&\mapsto \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\label{eq:rho:func}
\end{align}
is upper bounded almost surely by $\hat{\sigma}^k_+$.
In particular, if $\Lo_\mw$ is bounded below almost surely, \eqref{eq:rho:func} always satisfies assumption \eqref{assumpt:local} and Algorithm~\ref{alg:Ozaki:PaRIS} can be used. This condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.\ domains for which the exact algorithms EA1 and EA2 can be used.

@@

@@

>>>>

\section{The Generalized Random PaRIS algorithm}
\label{sec:rwparis}
%. These estimators are designed for stochastic processes
Let $(X_t)_{t\ge 0}$ be defined as a weak solution to the following SDE in $\mathbb{R}^d$:
\begin{equation}
\label{eq:target:sde}
X_0 = x_0\quad\mbox{and}\quad \rmd X_t = \alpha(X_t)\rmd t + \Gamma(X_t)\rmd W_t\eqsp,
\end{equation}
where $(W_t)_{t\ge 0}$ is a standard Brownian motion on $\mathbb{R}^d$, $\alpha: \mathbb{R}^d\to\mathbb{R}^d$ and $\Gamma: \mathbb{R}^d\to \mathbb{R}^{d\times d}$. 
In this paper, we focus on unbiased estimator of the unknown transition densities given by generalized Poisson estimators defined in \cite{fearnhead:papaspiliopoulos:roberts:2008}. The use of generalized Poisson estimators requires that the diffusion coefficient $\Gamma$ is constant and equals to the identity matrix. Note that, when this is not the case, there exists a class of diffusion processes that may be transformed to satisfy this assumption. This paper then focuses on reducible SDE for which there exists an invertible and infinitely differentiable function $\eta$ such that the process $\left\{Z_t = \eta(X_t)\right\}_{t\ge 0}$ satisfies the SDE  $z_0 = \eta(x_0)$ and
\begin{equation}
\label{eq:diff:lamperti}
\mathrm{d}Z_t = \beta(Z_t)\mathrm{d}t + \mathrm{d}W_t\eqsp.
\end{equation}
By Ito's formula,  it is straightforward to show that, in the case of a reducible diffusion, the Jacobian matrix of $\eta$ satisfies
\[
\nabla \eta = \Gamma^{-1}\eqsp,
\]
and, in the case $d=1$,
\[
\beta:u \mapsto \frac{\alpha(\eta^{-1}(u))}{\Gamma(\eta^{-1}(u))} - \frac{\Gamma'(\eta^{-1}(u))}{2}\eqsp.
\]
In the case of a scalar diffusion, this Lamperti transform is given by:
\[
\eta:x\mapsto \int_{x_0}^x\Gamma^{-1}(u)\mathrm{d}u\eqsp.
\]
In the general case, \cite[Proposition~$1$]{ait-sahalia:2008} shows that when $\Gamma$ is nonsingular, the SDE is reducible if and only if, for all $1\le i,j,k\le d$,
\begin{equation}
\label{eq:reducibility}
\frac{\partial\Gamma^{-1}_{i,j}}{\partial x_k} = \frac{\partial\Gamma^{-1}_{i,k}}{\partial x_j}\eqsp.
\end{equation}
In the case of a diagonal matrix $\Gamma$ \eqref{eq:reducibility} is equivalent to assume that $\Gamma$ is such that for each $1\le i\le d$, $\Gamma_{i,i}$ depends on $x_i$ only. \cite{ait-sahalia:2008} notes that the reducibility condition \eqref{eq:reducibility} holds also for some nondiagonal matrices $\Gamma$. This is true in particular in the case $d=2$ for stochastic volatility models where $\sigma$ is of the form:
\[
\Gamma(x) = \begin{pmatrix}a(x_1) & a(x_1)b(x_2)\\0&c(x_2)\end{pmatrix}\eqsp.
\]
In the following, it is assumed that the process $(X_t)_{t\ge 0}$ satisfies the SDE \eqref{eq:target:sde} with $\Gamma$ being the identity matrix, i.e. we consider a diffusion after the application of the Lamperti transform. In addition, designing general Poisson estimators also requires that:
\begin{enumerate}[i)]
\item $\alpha$ is of the form $\alpha(x) = \nabla_x A(x)$ where $A: \mathbb{R}^d \to \mathbb{R}$ is a twice continuously differentiable function ;
\item the function $x\mapsto \left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2$ is lower bounded where $\triangle$ is the Laplace operator.
\end{enumerate}
Assumption i) is the most restrictive \textcolor{red}{quelques exemples} while assumption ii) is satisfied in many situations \textcolor{red}{quelques exemples}. They are for instance satisfied for potential based movement models in ecology, see \cite{gloaguen:etienne:lecorff:2017}.

In the case where these assumptions are not fulfilled, an alternative to generalized Poisson estimators is given by continuous importance sampling procedures for SDE, see \cite{}.

The solution to \eqref{eq:target:sde} is supposed to be partially observed at times $t_0=0,\dots,t_n$ through an observation process $(Y_k)_{0\le k \le n}$ in $(\mathbb{R}^m)^{n+1}$. 
For all $0\le k \le n$, the distribution of $Y_k$ given $X_k:= X_{t_k}$ has a density with respect to a reference measure $\lambda$ on $\mathbb{R}^m$ given by $g(X_k,\cdot) = g_k(X_k)$. 
The distribution of $X_0$ has a density with respect to a reference measure $\mu$ on $\mathbb{R}^d$ given by $\chi$.
For all $0\le k \le n-1$, the conditional distribution of $X_{k+1} $ given $X_{k}$ has a density $\qk(X_{k},\cdot)$ with respect to $\mu$.

@@

\subsection*{Bounded estimator of $q_k$}
For $x, y \in \mathbb{R}^d$, by Girsanov and Ito's formulas, the transition density $q_k(x,y)$ of \eqref{eq:target:sde} satisfies, with $\Delta_k = t_{k+1}-t_k$,
\begin{align*}
q_k(x,y)=\varphi_{\Delta_k}(x,y)\exp\left\lbrace A(y)-A(x)\right\rbrace \mathbb{E}_{\mathbb{W}^{x,y,\Delta_k}}\left[ \exp \left\lbrace - \int_0^{\Delta_k} \phi(\mw_s)\rmd s \right\rbrace \right]\eqsp,
\end{align*}
where $\mathbb{W}^{x,y,\Delta_k}$ is the law of Brownian bridge starting at $x$ at 0 and hitting $y$ at $\Delta_k$, $(\mw_t)_{0\leq t \leq \Delta_k}$ is such a Brownian bridge, $\varphi_{\Delta_k}(x,y)$ is the p.d.f. of a normal distribution with mean $x$ and variance $\Delta_k$, evaluated at $y$ and $\phi:\mathbb{R}^d\to\mathbb{R}$ is defined as:
\[
\phi(x) =\left(\|\alpha(x)\|^2  + \triangle A(x)\right)/2\eqsp.
\]
Assume that there exist random variables $\Lo_\mw$ and $\U_\mw$ such that for all $0\leq s \leq \Delta_k$, $\Lo_\mw \leq \phi(\mw_s)\leq \U_\mw$.
Let $\kappa$ be a random variable taking values in $\mathbb{N}$ with distribution $\mu$ and $(U_j)_{1\le j\le \kappa}$ be independent uniform random variables on $[0,\Delta_k]$, and $
\zeta_k = \left\{\kappa,\mw,U_1,\ldots,U_\kappa\right\}\eqsp$. 
As shown in \cite{fearnhead:papaspiliopoulos:roberts:2008}, a positive unbiased estimator is given by 
\begin{multline}
\widehat{q}_k(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)\right\}\\ 
\times\mathrm{exp}\left\{-\U_\mw\Delta\right\}\frac{\Delta_k^{\kappa}}{\mu(\kappa)\kappa!}\prod_{j=1}^{\kappa}\left(\U_\mw-\phi(\mw_{U_j})\right)\eqsp.\label{eq:unbiased:q}
\end{multline}
Interesting choices of $\mu$ are discussed in \cite{fearnhead:papaspiliopoulos:roberts:2008} and we focus here on the so called GPE-1, where $\mu$ is a Poisson distribution with intensity $(\U_\mw-\Lo_\mw)\Delta_k$. In that case, the estimator \eqref{eq:unbiased:q} becomes:
\begin{equation}
\widehat{q}_{k}(x,y;\zeta_k) = \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\prod_{j=1}^{\kappa}\frac{\U_\mw-\phi(\mw_{U_j})}{\U_\mw-\Lo_\mw}\eqsp.\label{eq:GPE1}
\end{equation}
On the r.h.s. of \eqref{eq:GPE1}, the product over $\kappa$ elements is bounded by 1, therefore, a sufficient condition to satisfy of the assumptions \eqref{assumpt:global}-\eqref{assumpt:local} is that the function:
\begin{align}
\rho_{\Delta_k}:\eqsp\mathbb{R}^d\times \mathbb{R}^d &\mapsto \mathbb{R}\nonumber\\
(x,y)&\mapsto \varphi_{\Delta_k}(x,y) \exp \left\{A(y) - A(x)- \Lo_\mw\Delta_k \right\}\label{eq:rho:func}
\end{align}
is upper bounded almost surely by $\hat{\sigma}^k_+$.
In particular, if $\Lo_\mw$ is bounded below almost surely, \eqref{eq:rho:func} always satisfies assumption \eqref{assumpt:local} and Algorithm~\ref{alg:Ozaki:PaRIS} can be used. This condition is always satisfied for models in the domains $\mathcal{D}_1$ and $\mathcal{D}_2$ defined in \cite{beskos:papaspiliopoulos:roberts:fearnhead:2006}, i.e.\ domains for which the exact algorithms EA1 and EA2 can be used.

@@

Note that in this case the Lamperti transform to obtain a diffusion with a unitary diffusion term depends on $\sigma$. The process $(X_t)_{t\ge 0}$ is a function of $\sigma$ and is not directly observed if $\sigma$ is unknown, which prevents a direct use of an EM algorithm to estimate $\sigma$. Following \cite[Section~8.2]{beskos:papaspiliopoulos:roberts:fearnhead:2006}, this may be overcome with a two-step transformation of the process $(Z_t)_{t\ge 0}$. After the Lamperti transform, the process is transformed into a diffusion bridge  is used to obtai

@@

\bibitem{gloaguen:etienne:lecorff:2017}
P.~ Gloaguen, M.-P.~\'Etienne, and S.~Le~Corff.
\newblock Inference procedures to estimate potential based movement models in ecology.
\newblock {\em To appear in the {J}ournal of the {R}oyal {S}tatistical {S}ociety: {S}eries {C}}.
